Seasonal patterns across states

### 2017-02-15 temporal/seasonal-traces/seasonal-trace.Rmd
cirpo, doxy, etc.

---
title: "Seasonal patterns for azithromycin across states"
author: "Scott Olesen"
date: "2/15/2017"
output: html_document
---

@@

Azithromycin has a seasonal pattern in prescriptions. Does that pattern vary by geography? I used the Medicare data.

```{r load_data, echo=FALSE, message=FALSE}
dat2011 = read_tsv('../temporal-abx-2011.tsv')

full_dat = lapply(2011:2014, function(x) { read_tsv(sprintf('../temporal-abx-%i.tsv', x)) %>% mutate(year=x) }) %>%
  bind_rows()
```

@@

@@

```{r plot_azithro_two_states, echo=FALSE}
tick_labels = seq(1, 52) %>% { if_else(. %% 2 == 1, as.character(.), '') }
dat2011 %>%
  filter(abx == 'azithromycin', state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan'), week < 53.0) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=state, color=state)) +
  geom_point() + geom_line() +
  xlab('week of year 2011') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Azithromycin claims across states in 2011')
```

@@

```{r plot_azithro_years, echo=FALSE}
full_dat %>%
  filter(abx == 'azithromycin', state == 'Florida', week < 53.0) %>%
  mutate(year=factor(year)) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=year, color=year)) +
  geom_point() + geom_line() +
  xlab('week of year') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Azithromycin claims in Florida across years')
```

@@

full_dat %>%
  filter(abx == 'azithromycin', state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan', 'North Dakota', 'Mississippi'), week <= end_week) %>%
  group_by(year, state) %>%
  summarize(midpoint_time=midpoint_time(n_rx, week)) %>%
  mutate(midpoint_time=round(midpoint_time, 1)) %>%
  spread(year, midpoint_time) %>%
  arrange(desc(`2011`)) %>%
  kable
```

>>>>

---
title: "Seasonal patterns across states"
author: "Scott Olesen"
date: "2/15/2017"
output: html_document
---

@@

Azithromycin has a seasonal pattern in prescriptions. Does that pattern vary by geography? What about for other drugs, like ciprofloxacin or tetracycline?

@@

I used the Medicare data.

```{r load_data, echo=FALSE, message=FALSE}
dat = lapply(2011:2014, function(x) { read_tsv(sprintf('../temporal-abx-%i.tsv', x)) %>% mutate(year=x) }) %>%
  bind_rows()
```

# Results

## Azithromycin

@@

```{r plot_azithro_two_states, echo=FALSE}
tick_labels = seq(1, 52) %>% { if_else(. %% 2 == 1, as.character(.), '') }
dat %>%
  filter(year == 2011) %>%
  filter(abx == 'azithromycin', state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan'), week < 53.0) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=state, color=state)) +
  geom_point() + geom_line() +
  xlab('week of year 2011') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Azithromycin claims across states in 2011')
```

@@

```{r plot_azithro_years, echo=FALSE}
dat %>%
  filter(abx == 'azithromycin', state == 'Florida', week < 53.0) %>%
  mutate(year=factor(year)) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=year, color=year)) +
  geom_point() + geom_line() +
  xlab('week of year') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Azithromycin claims in Florida across years')
```

@@

dat %>%
  filter(abx == 'azithromycin', state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan', 'North Dakota', 'Mississippi'), week <= end_week) %>%
  group_by(year, state) %>%
  summarize(midpoint_time=midpoint_time(n_rx, week)) %>%
  mutate(midpoint_time=round(midpoint_time, 1)) %>%
  spread(year, midpoint_time) %>%
  arrange(desc(`2011`)) %>%
  kable
```

## Tetracycline

There are multiple drugs with the substring "cycline". I grouped the consumption of these together for this analysis.

```{r tetracycline_names, echo=FALSE}
library(stringr)
tetracyclines = dat$abx %>% unique %>% { .[str_detect(., 'cycline')] }
tetracyclines
```

Tetracycline does appear to have some seasonal patterns, but they apparently differ by states: in California, Florida, and Texas (southern states), they are prescribed more in winter than in summer (although the seasonal variation seems to be around half as large as for azithromycin). In New York and Michigan (northern states), tetracycline claims appear to peak in the fall.

```{r tetracycline_seasonality, echo=FALSE}
dat %>%
  filter(year == 2011) %>%
  # group all the tetracyclines together
  filter(abx %in% tetracyclines) %>%
  group_by(week, state) %>%
  summarize(n_rx=sum(n_rx)) %>%
  filter(state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan'), week < 53.0) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=state, color=state)) +
  geom_point() + geom_line() +
  xlab('week of year 2011') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Tetracycline (etc.) claims across states in 2011')
```

## Ciprofloxacin

For comparison, I also reproduced the plots for ciprofloxacin claims. These appear flat.

```{r cipro_seasonality, echo=FALSE}
dat %>%
  filter(year == 2011) %>%
  filter(abx=='ciprofloxacin', state %in% c('Florida', 'New York', 'Texas', 'California', 'Michigan'), week < 53.0) %>%
  ggplot(aes(x=factor(week), y=n_rx, group=state, color=state)) +
  geom_point() + geom_line() +
  xlab('week of year 2011') + ylab('# of claims') +
  scale_x_discrete(labels=tick_labels) +
  theme_minimal() +
  ggtitle('Cipro claims across states in 2011')
```
